---
title: "scRNA-seq of human PBMCs treated with mechanically generated extracellular vesicles (mEV)"
author: "Keng Jung Lee (Zheng Lab)"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview

Single-cell RNA-seq analysis of human peripheral blood mononuclear cells (PBMCs)
treated *ex vivo* with **mechanically generated EVs (mEV)** shed by circulating
tumor cells arrested in the microvasculature, **canonical secreted EVs (sEV)**,
or **PBS** (vehicle control). Four donors (606, 610, 706, 746) were each split
across the three conditions and processed with cell hashing (HTO) on the 10x
Genomics platform.

mEVs are enriched in the nuclear DAMP HMGB1 and potently activate monocytes;
this experiment profiles the single-cell PBMC response to mEV vs sEV, focusing
on the monocyte and dendritic-cell compartments.

| Sample | Donor | Condition | HTO multiplexed |
|--------|-------|-----------|-----------------|
| 606M / 610M / 706M / 746M | 606/610/706/746 | mEV | yes |
| 606P / 610P / 706P / 746P | 606/610/706/746 | PBS | yes |
| 606S / 610S / 706S / 746S | 606/610/706/746 | sEV | yes |

**Pipeline:** QC → HTO demultiplexing → filtering → SCTransform → CCA
integration → clustering → marker-based annotation → pseudobulk DESeq2 DE
(mEV vs sEV) → GO enrichment → Azimuth reference mapping.

> **Paths:** raw `cellranger` outputs are expected under `data/post_aggr/<sample>/`.
> Edit the `data_dir` variable below to point at your local copy.

# Setup

```{r libraries}
library(Seurat)
library(Matrix)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(ggridges)
library(patchwork)
library(monocle3)
```

```{r config}
# Root directory holding cellranger filtered matrices, one folder per sample.
data_dir <- "data/post_aggr"

samples <- c("606M","606P","606S",
             "610M","610P","610S",
             "706M","706P","706S",
             "746M","746P","746S")

# Map sample suffix -> treatment condition.
suffix_to_condition <- c(M = "mEV", P = "PBS", S = "sEV")
```

# 1. Load 10x data

Each sample contains both a Gene Expression and an Antibody Capture (HTO) assay.

```{r load-10x}
raw <- lapply(samples, function(s) {
  Read10X(data.dir = file.path(data_dir, s, "outs/count/filtered_feature_bc_matrix"))
})
names(raw) <- samples
```

# 2. Create Seurat objects (RNA + HTO) and add metadata

```{r create-objects}
objects <- lapply(samples, function(s) {
  obj <- CreateSeuratObject(counts = raw[[s]][["Gene Expression"]])
  obj[["hto"]] <- CreateAssayObject(counts = raw[[s]][["Antibody Capture"]])

  condition <- suffix_to_condition[substr(s, 4, 4)]   # last char: M/P/S
  obj <- AddMetaData(obj, metadata = condition, col.name = "Condition")
  obj <- AddMetaData(obj, metadata = s,          col.name = "Sample")
  obj
})
names(objects) <- samples
```

# 3. QC metrics

```{r qc-metrics}
objects <- lapply(objects, function(obj) {
  obj[["percent.mt"]] <- PercentageFeatureSet(obj, pattern = "^MT-")
  obj
})

# Per-sample QC visualisation
for (s in samples) {
  print(
    VlnPlot(objects[[s]],
            features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
            ncol = 3) + patchwork::plot_annotation(title = s)
  )
  print(
    FeatureScatter(objects[[s]], feature1 = "nCount_RNA", feature2 = "nFeature_RNA") +
      geom_smooth() + ggtitle(s)
  )
}
```

# 4. Merge

```{r merge}
merged <- merge(
  objects[[1]], y = objects[-1],
  add.cell.ids = samples,
  project = "merged_pbmc"
)
merged
```

# 5. HTO demultiplexing

```{r hto-demux}
# Log-normalize RNA; CLR-normalize HTO.
merged <- NormalizeData(merged, assay = "RNA")
merged <- FindVariableFeatures(merged, selection.method = "mean.var.plot", assay = "RNA")
merged <- ScaleData(merged, features = VariableFeatures(merged), assay = "RNA")

merged <- NormalizeData(merged, assay = "hto", normalization.method = "CLR")
merged <- HTODemux(merged, assay = "hto", positive.quantile = 0.99)

table(merged$hto_classification.global)
```

```{r hto-plots, fig.width=6, fig.height=5}
Idents(merged) <- "hto_maxID"
RidgePlot(merged, assay = "hto", features = rownames(merged[["hto"]])[1:2], ncol = 2)
FeatureScatter(merged, feature1 = "Hu.Hashtag-1", feature2 = "Hu.Hashtag-2")
```

```{r hto-tsne}
# Drop HTO-negative droplets and visualise singlet/doublet structure on an
# HTO-distance tSNE.
Idents(merged) <- "hto_classification.global"
merged.subset <- subset(merged, idents = "Negative", invert = TRUE)

hto.dist.mtx <- as.matrix(dist(t(GetAssayData(merged.subset, assay = "hto"))))
merged.subset <- RunTSNE(merged.subset, distance.matrix = hto.dist.mtx, perplexity = 100)
DimPlot(merged.subset)
```

# 6. QC filtering

```{r filter}
# Retain cells with 200 < nFeature_RNA < 8000 and < 5% mitochondrial reads.
merged <- subset(merged, subset = nFeature_RNA > 200 & nFeature_RNA < 8000 & percent.mt < 5)
```

# 7. Normalization, PCA, UMAP

```{r sctransform}
merged <- SCTransform(merged)
merged <- RunPCA(merged)
merged <- RunUMAP(merged, dims = 1:30)

DimPlot(merged, reduction = "umap", group.by = c("Condition", "orig.ident"))
```

# 8. Integration (CCA) and clustering

```{r integrate}
merged_integ <- IntegrateLayers(
  object = merged, method = CCAIntegration,
  normalization.method = "SCT", verbose = FALSE
)
merged_integ <- FindNeighbors(merged_integ, reduction = "integrated.dr", dims = 1:10)
merged_integ <- FindClusters(merged_integ, resolution = 0.8)
merged_integ <- RunUMAP(merged_integ, dims = 1:10, reduction = "integrated.dr")

DimPlot(merged_integ, reduction = "umap", label = TRUE, label.size = 3, alpha = 0.5)
DimPlot(merged_integ, reduction = "umap", split.by = "Condition",
        label = TRUE, label.size = 3, alpha = 0.5)
DimPlot(merged_integ, reduction = "umap", split.by = "Sample",
        label = TRUE, label.size = 3, alpha = 0.5)
```

# 9. Canonical marker expression

```{r marker-panel}
# Lineage markers used to guide cluster annotation.
markers <- c("PTPRC","MS4A1","GNLY","CD3D","CD3E","CD3G","CD14","CD68",
             "FCER1A","FCGR3A","LYZ","PPBP","CD8A","GZMB","CD4","CD79A",
             "CD79B","HBG1","HBG2","KLRF1","XCL1")
FeaturePlot(merged_integ, features = markers, pt.size = 0.5)
```

```{r marker-by-condition}
# Markers of interest, split by treatment condition.
markers_by_condition <- c("CD3E","CD3G","CD8A","CD8B","MS4A1","CD79A","CD79B",
                          "CD19","CD14","LYZ","FCER1A","CD4","IL7R","CCR7",
                          "SELL","S100A4","FCGR3A","NKG7","GNLY","SLC4A10",
                          "CST3","IL3RA","CD27","MKI67","GSDMD","CD271")
for (g in markers_by_condition) {
  print(FeaturePlot(merged_integ, features = g, pt.size = 0.5, split.by = "Condition"))
}
```

# 10. Cluster markers

```{r find-markers}
merged_integ_prep <- PrepSCTFindMarkers(merged_integ, assay = "SCT", verbose = TRUE)
merged_integ.markers <- FindAllMarkers(merged_integ_prep)

merged_integ.markers %>%
  group_by(cluster) %>%
  dplyr::filter(avg_log2FC > 1) %>%
  slice_head(n = 10) %>%
  ungroup() -> top10

DoHeatmap(merged_integ, features = top10$gene) + NoLegend()
```

# 11. Cell-type annotation

```{r annotate}
# Cluster -> cell-type assignment based on canonical markers above.
labeled.merged_integ <- RenameIdents(
  merged_integ,
  `0` = "Memory CD4 T", `1` = "Naive CD4 T", `2` = "Naive CD4 T",
  `3` = "Monocytes",    `4` = "NK",          `5` = "NK",
  `6` = "Monocytes",    `7` = "CD8 T",       `8` = "CD8 T",
  `9` = "CD8 T",        `10` = "Naive CD4 T",`11` = "MAIT T",
  `12` = "B",           `13` = "Naive CD4 T",`14` = "DC",
  `15` = "NK T",        `16` = "B",          `17` = "Memory CD4 T",
  `18` = "Naive CD4 T", `19` = "Naive CD4 T"
)
DimPlot(labeled.merged_integ, label = TRUE)
DimPlot(labeled.merged_integ, label = TRUE, split.by = "Condition")
```

# 12. Pseudobulk differential expression (mEV vs sEV)

Aggregate to individual × cell-type pseudobulk profiles, then run DESeq2 within
a cell type comparing mEV vs sEV.

```{r pseudobulk}
bulk <- AggregateExpression(
  labeled.merged_integ, return.seurat = TRUE, slot = "counts",
  assays = "RNA", group.by = c("ident", "Sample", "Condition")
)
```

```{r de-monocytes}
mono.bulk <- subset(bulk, orig.ident == "Monocytes")
Idents(mono.bulk) <- "Condition"
de_markers <- FindMarkers(mono.bulk, ident.1 = "mEV", ident.2 = "sEV",
                          slot = "counts", test.use = "DESeq2", verbose = FALSE)
de_markers$gene <- rownames(de_markers)

library(ggrepel)
ggplot(de_markers, aes(avg_log2FC, -log10(p_val))) +
  geom_point(size = 0.5, alpha = 0.5) + theme_bw() +
  ylab("-log10(unadjusted p-value)") +
  geom_text_repel(aes(label = ifelse(p_val_adj < 0.05, gene, "")),
                  colour = "red", size = 3)

write.csv(de_markers, file = "results/mono_deg_mEVvssEV.csv")
```

```{r de-dc}
DC.bulk <- subset(bulk, orig.ident == "DC")
Idents(DC.bulk) <- "Condition"
de_markers_DC <- FindMarkers(DC.bulk, ident.1 = "mEV", ident.2 = "sEV",
                             slot = "counts", test.use = "DESeq2", verbose = FALSE)
de_markers_DC$gene <- rownames(de_markers_DC)

ggplot(de_markers_DC, aes(avg_log2FC, -log10(p_val))) +
  geom_point(size = 0.5, alpha = 0.5) + theme_bw() +
  ylab("-log10(unadjusted p-value)") +
  geom_text_repel(aes(label = ifelse(p_val_adj < 0.05, gene, "")),
                  colour = "red", size = 3)
```

# 13. GO enrichment of mEV-upregulated genes (monocytes)

```{r go}
library(clusterProfiler)
library(org.Hs.eg.db)
library(AnnotationDbi)

up_gene_mEVvssEV <- rownames(de_markers[de_markers$avg_log2FC > 0, ])

GO_results <- enrichGO(
  gene    = up_gene_mEVvssEV,
  OrgDb   = org.Hs.eg.db,
  keyType = "SYMBOL",
  ont     = "BP"
)
as.data.frame(GO_results)
barplot(GO_results, showCategory = 20)
```

# 14. Azimuth reference mapping

```{r azimuth}
library(Azimuth)
library(SeuratData)

options(timeout = 1000)
merged.mapped <- RunAzimuth(merged, reference = "pbmcref")

p1 <- DimPlot(merged.mapped, group.by = "predicted.celltype.l2",
              label = TRUE, label.size = 3) + NoLegend()
p2 <- DimPlot(merged.mapped, group.by = "Condition")
p1 + p2
```

# Session info

```{r session-info}
sessionInfo()
```
