---
title: "scRNA-seq of human PBMCs treated with mEV — Azimuth reference mapping and CD14 monocyte analysis"
author: "Keng Jung Lee (Zheng Lab)"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview

Companion notebook to `mEV-PBMC.Rmd`. Instead of HTO demultiplexing + CCA
integration + manual annotation, this notebook annotates the same 12 PBMC
libraries by **Azimuth reference mapping** to the `pbmcref` PBMC atlas, then
focuses downstream analysis on the **CD14 monocyte** compartment — the cell type
most strongly activated by mechanically generated EVs (mEV) via the HMGB1 axis.

Single-cell RNA-seq of human PBMCs treated *ex vivo* with **mechanically
generated EVs (mEV)**, **canonical secreted EVs (sEV)**, or **PBS** (vehicle).
Four donors (606, 610, 706, 746) were each split across the three conditions on
the 10x Genomics platform.

**Pipeline:** load → QC → merge → filter → normalize → Azimuth mapping
(`pbmcref`, `predicted.celltype.l2`) → CD14-monocyte pairwise DE across the three
conditions → GO/GSEA enrichment → optional CD14 subclustering and trajectory
analysis.

> **Paths:** raw `cellranger` outputs are expected under
> `data/post_aggr/<sample>/`. Edit the `data_dir` variable below to point at your
> local copy. Tables and figures are written to `results/`.

# Setup

```{r libraries}
library(Seurat)
library(Matrix)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(ggridges)
library(ggrepel)
library(patchwork)
library(data.table)
```

```{r config}
# Root directory holding cellranger filtered matrices, one folder per sample.
data_dir    <- "data/post_aggr"
results_dir <- "results"
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)

samples <- c("606M","606P","606S",
             "610M","610P","610S",
             "706M","706P","706S",
             "746M","746P","746S")

# Map sample suffix (last character) -> treatment condition.
suffix_to_condition <- c(M = "mEV", P = "PBS", S = "sEV")
```

# 1. Load 10x data

```{r load-10x}
raw <- lapply(samples, function(s) {
  Read10X(data.dir = file.path(data_dir, s, "outs/count/filtered_feature_bc_matrix"))
})
names(raw) <- samples
```

# 2. Create Seurat objects and add metadata

```{r create-objects}
objects <- lapply(samples, function(s) {
  counts <- raw[[s]]
  # Multi-assay runs return a list; take the Gene Expression matrix.
  if (is.list(counts)) counts <- counts[["Gene Expression"]]

  obj <- CreateSeuratObject(counts = counts)
  condition <- suffix_to_condition[substr(s, 4, 4)]
  obj <- AddMetaData(obj, metadata = condition, col.name = "Condition")
  obj <- AddMetaData(obj, metadata = s,          col.name = "Sample")
  obj
})
names(objects) <- samples
```

# 3. QC metrics

```{r qc-metrics}
objects <- lapply(objects, function(obj) {
  obj[["percent.mt"]] <- PercentageFeatureSet(obj, pattern = "^MT-")
  obj
})

# Per-sample QC visualisation.
for (s in samples) {
  print(
    VlnPlot(objects[[s]],
            features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
            ncol = 3) + patchwork::plot_annotation(title = s)
  )
  print(
    FeatureScatter(objects[[s]], feature1 = "nCount_RNA", feature2 = "nFeature_RNA") +
      geom_smooth() + ggtitle(s)
  )
}
```

# 4. Merge

```{r merge}
merged <- merge(
  objects[[1]], y = objects[-1],
  add.cell.ids = samples,
  project = "merged_pbmc"
)
merged
```

# 5. QC filtering

```{r filter}
# Retain cells with 200 < nFeature_RNA < 8000 and < 5% mitochondrial reads.
# (Filtering is applied to the merged object so that downstream steps operate on
# the filtered cells — in the original exploratory notebook the per-sample
# subset() calls ran *after* the merge and were therefore never propagated.)
merged <- subset(merged, subset = nFeature_RNA > 200 & nFeature_RNA < 8000 & percent.mt < 5)
merged
```

# 6. Normalization

```{r normalize}
merged <- NormalizeData(merged, assay = "RNA")
merged <- FindVariableFeatures(merged, selection.method = "mean.var.plot", assay = "RNA")
merged <- ScaleData(merged, features = VariableFeatures(merged), assay = "RNA")
```

# 7. Azimuth reference mapping

`RunAzimuth` is slow and downloads the `pbmcref` reference at runtime. Run the
compute chunk once, save the result, and reload from cache on subsequent knits.

```{r azimuth-run, eval=FALSE}
library(Azimuth)
library(SeuratData)

options(timeout = 1000)
merged.mapped <- RunAzimuth(merged, reference = "pbmcref")
saveRDS(merged.mapped, file.path(results_dir, "merged.mapped.rds"))
```

```{r azimuth-load}
# Reload the cached mapping produced by the chunk above.
merged.mapped <- readRDS(file.path(results_dir, "merged.mapped.rds"))
```

```{r azimuth-dimplots, fig.width=20, fig.height=10}
# UMAP coloured by predicted cell type (l2), split by treatment condition.
DimPlot(merged.mapped, group.by = "predicted.celltype.l2",
        split.by = "Condition", label = TRUE, label.size = 3,
        repel = TRUE, alpha = 0.3)
```

```{r azimuth-dimplot-all, fig.width=12, fig.height=10}
DimPlot(merged.mapped, group.by = "predicted.celltype.l2",
        label = TRUE, label.size = 5, repel = TRUE, alpha = 0.2)
```

```{r azimuth-score, fig.width=12, fig.height=8}
# Mapping-confidence score per predicted cell type.
VlnPlot(merged.mapped, features = "predicted.celltype.l2.score")
```

# 8. Cell-type composition

```{r composition}
md <- as.data.table(merged.mapped@meta.data)
celltype_counts <- md[, .N, by = c("Condition", "predicted.celltype.l2")]
celltype_counts <- celltype_counts[order(Condition, -N)]
celltype_counts

write.csv(celltype_counts, file.path(results_dir, "celltype_counts_by_condition.csv"),
          row.names = FALSE)
```

# 9. Marker expression in monocytes

Inflammatory, pyroptosis, and tissue-factor / thrombosis markers central to the
mEV–HMGB1–monocyte axis.

```{r prep-de}
# SCTransform for marker visualisation and within-cell-type DE.
merged.mapped <- SCTransform(merged.mapped)
Idents(merged.mapped) <- "predicted.celltype.l2"
```

```{r marker-vln}
VlnPlot(merged.mapped, features = c("GSDMD"),
        split.by = "Condition", group.by = "predicted.celltype.l2",
        idents = "CD14 Mono")

VlnPlot(merged.mapped, features = c("TNF", "IL1B", "IL6"),
        split.by = "Condition", group.by = "predicted.celltype.l2",
        idents = "CD16 Mono")
```

```{r marker-feature, fig.width=12, fig.height=5}
# Tissue factor (F3) and CD14, by condition.
FeaturePlot(merged.mapped, features = c("F3", "CD14"), split.by = "Condition")
```

# 10. CD14 monocyte differential expression (pairwise across conditions)

Within CD14 monocytes, compare each pair of conditions on the corrected SCT
counts. `PrepSCTFindMarkers` recomputes the corrected counts after subsetting.

```{r de-prep}
merged.mapped.prepfind <- PrepSCTFindMarkers(merged.mapped)
merged.mapped.prepfind$celltype.condition <-
  paste(merged.mapped.prepfind$predicted.celltype.l2,
        merged.mapped.prepfind$Condition, sep = "_")
Idents(merged.mapped.prepfind) <- "celltype.condition"
```

```{r de-run}
# Helper: FindMarkers between two CD14-monocyte condition groups.
cd14_de <- function(cond1, cond2) {
  res <- FindMarkers(merged.mapped.prepfind,
                     ident.1 = paste0("CD14 Mono_", cond1),
                     ident.2 = paste0("CD14 Mono_", cond2),
                     verbose = FALSE)
  res$gene <- rownames(res)
  res
}

de_markers_mvss <- cd14_de("mEV", "sEV")   # mEV vs sEV
de_markers_mvsp <- cd14_de("mEV", "PBS")   # mEV vs PBS
de_markers_svsp <- cd14_de("sEV", "PBS")   # sEV vs PBS

write.csv(de_markers_mvss, file.path(results_dir, "de_markers_mvss_cd14.csv"))
write.csv(de_markers_mvsp, file.path(results_dir, "de_markers_mvsp_cd14.csv"))
write.csv(de_markers_svsp, file.path(results_dir, "de_markers_svsp_cd14.csv"))
```

```{r volcano}
# Helper: volcano plot labelling genes past adjusted-p and fold-change cutoffs.
volcano <- function(de, title, padj = 0.05, lfc = 1) {
  ggplot(de, aes(avg_log2FC, -log10(p_val))) +
    geom_point(size = 0.3, alpha = 0.3) +
    theme_bw() +
    ylab("-log10(unadjusted p-value)") +
    ggtitle(title) +
    geom_text_repel(
      aes(label = ifelse(p_val_adj < padj & avg_log2FC > lfc, gene, "")),
      colour = "red", size = 3)
}

volcano(de_markers_mvss, "CD14 Mono: mEV vs sEV")
volcano(de_markers_mvsp, "CD14 Mono: mEV vs PBS")
volcano(de_markers_svsp, "CD14 Mono: sEV vs PBS")
```

```{r cellscatter, fig.width=10, fig.height=5}
library(cowplot)
theme_set(theme_cowplot())
genes.to.label <- c("IL1B", "S100A9", "IFI6", "ISG20", "MX1",
                    "IFIT2", "IFIT1", "CXCL10", "CCL8")
p1 <- CellScatter(merged.mapped.prepfind, "CD14 Mono_PBS", "CD14 Mono_mEV",
                  highlight = genes.to.label)
LabelPoints(plot = p1, points = genes.to.label, repel = TRUE)
```

# 11. GO enrichment (GSEA) of CD14 DE contrasts

```{r gsea-libs}
library(clusterProfiler)
library(org.Hs.eg.db)
require(DOSE)
```

```{r gsea-run}
# Helper: ranked gene list (by avg_log2FC) -> gseGO over Biological Process.
run_gse <- function(de) {
  de <- de[order(de$avg_log2FC, decreasing = TRUE), ]
  gl <- de$avg_log2FC
  names(gl) <- rownames(de)
  gseGO(gl, ont = "BP", keyType = "SYMBOL",
        OrgDb = "org.Hs.eg.db", eps = 1e-300)
}

gse.mvss.cd14 <- run_gse(de_markers_mvss)
gse.mvsp.cd14 <- run_gse(de_markers_mvsp)
gse.svsp.cd14 <- run_gse(de_markers_svsp)

write.csv(as.data.frame(gse.mvss.cd14), file.path(results_dir, "gse.mvss.cd14.csv"))
write.csv(as.data.frame(gse.mvsp.cd14), file.path(results_dir, "gse.mvsp.cd14.csv"))
write.csv(as.data.frame(gse.svsp.cd14), file.path(results_dir, "gse.svsp.cd14.csv"))
```

```{r gsea-dotplots, fig.width=12, fig.height=18}
dotplot(gse.mvss.cd14, showCategory = 10, split = ".sign") + facet_grid(. ~ .sign)
dotplot(gse.mvsp.cd14, showCategory = 10, split = ".sign") + facet_grid(. ~ .sign)
dotplot(gse.svsp.cd14, showCategory = 10, split = ".sign") + facet_grid(. ~ .sign)
```

```{r gsea-curves}
# Representative enrichment curves (edit geneSetID to a term in your results).
# NF-kB signalling (mEV vs PBS):
gseaplot(gse.mvsp.cd14, geneSetID = "GO:0007249")
# Pattern-recognition / NLRP3-related terms (mEV vs sEV):
gseaplot(gse.mvss.cd14, geneSetID = "GO:0062208")
```

# 12. CD14 monocyte subclustering (optional)

Re-embed CD14 monocytes alone to resolve activation substructure.

```{r cd14-subset}
cd14 <- subset(merged.mapped, idents = "CD14 Mono")
cd14 <- NormalizeData(cd14)
cd14 <- FindVariableFeatures(cd14, nfeatures = 3000, assay = "RNA")
cd14 <- ScaleData(cd14)
cd14 <- RunPCA(cd14)
cd14 <- FindNeighbors(cd14)
cd14 <- FindClusters(cd14, resolution = 0.4)
cd14 <- RunUMAP(cd14, n.neighbors = 10, dims = 1:50, spread = 2, min.dist = 0.3)
```

```{r cd14-dimplots, fig.width=10, fig.height=5}
# Okabe-Ito colour-blind-safe palette.
okabe_ito <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442",
               "#0072B2", "#D55E00", "#CC79A7", "#999999")

DimPlot(cd14, group.by = "seurat_clusters", repel = TRUE,
        alpha = 0.7, cols = okabe_ito)
DimPlot(cd14, group.by = "seurat_clusters", split.by = "Condition",
        label = TRUE, repel = TRUE, alpha = 0.7, cols = okabe_ito)
DimPlot(cd14, group.by = "Condition", alpha = 0.7)
```

```{r cd14-features, fig.width=12, fig.height=5}
FeaturePlot(cd14, features = c("TNF", "IL1B", "HMGB1"), split.by = "Condition")
FeaturePlot(cd14, features = "PLAUR", split.by = "Condition")
FeaturePlot(cd14, features = "CXCL10", split.by = "Condition")

VlnPlot(cd14, features = c("PGK1", "SERPINE1", "S100A9", "RUNX1",
                           "PLAUR", "PLGRKT", "HPN"))
```

# 13. Combined monocyte (CD14 + CD16) embedding (optional)

```{r mono-subset}
mono <- subset(merged.mapped, idents = c("CD14 Mono", "CD16 Mono"))
mono <- NormalizeData(mono)
mono <- RunPCA(mono)
mono <- FindNeighbors(mono)
mono <- FindClusters(mono, resolution = 1)
mono <- RunUMAP(mono, n.neighbors = 10, dims = 1:50, spread = 2, min.dist = 0.3)

DimPlot(mono, reduction = "umap", split.by = "Condition",
        group.by = "predicted.celltype.l2")
DimPlot(mono, reduction = "umap")
```

# 14. Trajectory analysis (exploratory)

Slingshot lineage inference over the CD14 monocyte subclusters. This is
exploratory: the starting cluster should be chosen from the biology (e.g. the
least-activated resting cluster), so the chunk is not evaluated by default.

```{r slingshot, eval=FALSE}
library(slingshot)
library(SingleCellExperiment)

set.seed(1)
dimred     <- Embeddings(cd14, "umap")
clustering <- cd14$seurat_clusters

# Fit lineages and principal curves (set start.clus to your resting cluster).
sds <- slingshot(dimred, clusterLabels = clustering, stretch = 0)

# Colour cells by cluster and overlay the inferred lineage(s).
pal <- setNames(okabe_ito[seq_along(levels(clustering))], levels(clustering))
plot(dimred, col = pal[clustering], pch = 16, cex = 0.5, asp = 1,
     xlab = "UMAP1", ylab = "UMAP2", main = "Slingshot trajectory")
lines(SlingshotDataSet(sds), lwd = 2, col = "black")
```

# Appendix: mouse macrophage marker enrichment (external data, optional)

Side analysis on an external mouse macrophage marker table
(`data/macmarkers.csv`); unrelated to the human PBMC pipeline above. Provided
for reference only.

```{r mac-gsea, eval=FALSE}
library(org.Mm.eg.db)

mac <- read.csv(file.path(data_dir, "..", "macmarkers.csv"), header = TRUE)
mac <- mac[order(mac$avg_log2FC, decreasing = TRUE), ]
gl_mac <- mac$avg_log2FC
names(gl_mac) <- mac$gene

gse.mac <- gseGO(gl_mac, ont = "BP", keyType = "SYMBOL",
                 OrgDb = "org.Mm.eg.db", eps = 1e-300)
dotplot(gse.mac, showCategory = 7, split = ".sign") + facet_grid(. ~ .sign)
```

# Session info

```{r session-info}
sessionInfo()
```
